---
title: "Performance Analysis of Bloom Filter Joins"
author: "Gaurav Sood"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Analysis of Bloom Filter Joins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette compares `bloomjoin` against standard `dplyr` joins on two dimensions: **speed** and **memory**. Bloom joins excel when tables are asymmetric (large left, small right) with low overlap.

```{r setup, message = FALSE}
library(bloomjoin)
library(dplyr)
library(bench)
library(knitr)
```

```{r helpers, include = FALSE}
generate_data <- function(n_x, n_y, overlap, seed = 123) {
  set.seed(seed)
  y_keys <- seq_len(n_y)
  n_match <- round(n_x * overlap)
  x_matching <- if (n_match > 0 && n_match <= n_y) {
    sample(y_keys, n_match)
  } else if (n_match > n_y) {
    sample(y_keys, n_match, replace = TRUE)
  } else {
    integer(0)
  }
  x_nonmatching <- seq(n_y + 1, n_y + n_x - length(x_matching))

  list(
    x = data.frame(id = sample(c(x_matching, x_nonmatching)), xv = rnorm(n_x)),
    y = data.frame(id = y_keys, yv = rnorm(n_y))
  )
}

run_bench <- function(n_x, n_y, overlap, reps = 3) {
  dat <- generate_data(n_x, n_y, overlap)

  # Warm up

  invisible(bloom_join(dat$x, dat$y, by = "id"))
  invisible(inner_join(dat$x, dat$y, by = "id"))

  result <- bench::mark(
    bloom = bloom_join(dat$x, dat$y, by = "id"),
    dplyr = inner_join(dat$x, dat$y, by = "id"),
    iterations = reps,
    check = FALSE,
    memory = TRUE
  )

  meta <- attr(bloom_join(dat$x, dat$y, by = "id"), "bloom_metadata")

  data.frame(
    n_x = format(n_x, big.mark = ",", scientific = FALSE),
    n_y = format(n_y, big.mark = ",", scientific = FALSE),
    overlap = paste0(overlap * 100, "%"),
    speed_ratio = round(as.numeric(result$median[2]) / as.numeric(result$median[1]), 2),
    mem_ratio = round(as.numeric(result$mem_alloc[2]) / as.numeric(result$mem_alloc[1]), 2),
    reduction = paste0(round(meta$reduction_ratio * 100, 0), "%")
  )
}
```

## Correctness

First, verify that `bloom_join()` produces identical results to `dplyr`:

```{r correctness}
dat <- generate_data(5000, 500, 0.1)
bloom_result <- bloom_join(dat$x, dat$y, by = "id") |> arrange(id)
dplyr_result <- inner_join(dat$x, dat$y, by = "id") |> arrange(id)

# Strip bloom metadata for comparison
bloom_cmp <- as.data.frame(bloom_result)
attr(bloom_cmp, "bloom_metadata") <- NULL

all.equal(bloom_cmp, as.data.frame(dplyr_result), check.attributes = FALSE)
```

## Speed and Memory Benchmarks

We test across scenarios varying table asymmetry and overlap. `speed_ratio` and `mem_ratio` > 1 means bloom wins.

```{r benchmarks, message = FALSE, warning = FALSE}
scenarios <- list(
  c(1e6, 1e4, 0.01),
  c(1e6, 1e4, 0.05),
  c(5e5, 5e3, 0.02),
  c(5e5, 5e3, 0.10),
  c(2e5, 2e4, 0.05),
  c(2e5, 2e4, 0.25),
  c(1e5, 1e5, 0.10),
  c(1e5, 1e5, 0.50)
)

results <- do.call(rbind, lapply(scenarios, function(s) {
  run_bench(s[1], s[2], s[3], reps = 3)
}))

kable(results,
      col.names = c("n_x", "n_y", "overlap", "speed", "memory", "reduction"),
      align = "rrrrrr")
```

## Interpretation

- **speed, memory > 1**: bloom wins (dplyr_time / bloom_time)
- **reduction**: % of rows filtered out before join — explains memory savings

High reduction means fewer rows held in memory during the join. When reduction is low, Bloom filter overhead dominates and dplyr wins.

## When to Use Bloom Joins

| Condition | Bloom benefit |
|-----------|---------------|
| Large x, small y (10:1+) | Strong |
| Low overlap (<25%) | Strong |
| Equal-sized tables | None (use dplyr) |
| High overlap (>50%) | None (use dplyr) |

## Tuning

The `fpr` parameter controls the false positive rate. Lower values reduce false positives but increase filter size:

```{r tuning, message = FALSE}
dat <- generate_data(50000, 5000, 0.05)

fpr_results <- do.call(rbind, lapply(c(0.001, 0.01, 0.1), function(fpr) {
  result <- bench::mark(
    bloom_join(dat$x, dat$y, by = "id", fpr = fpr),
    iterations = 3,
    check = FALSE
  )
  data.frame(
    fpr = fpr,
    median_ms = round(as.numeric(result$median) * 1000, 1),
    mem_mb = round(as.numeric(result$mem_alloc) / 1024^2, 1)
  )
}))

kable(fpr_results, col.names = c("FPR", "Time (ms)", "Memory (MB)"))
```

Default `fpr = 0.01` balances speed and memory well.
